---
title: "Temperature in Te Ika-a-Māui / North Island over time"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    css: "body {\n  max-width: 100%;\n}\n"
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

Submitted by:

-   Arman Choudhry (300574050)

-   Amanda Best (300630839)

## Background and Data

Data for this project was obtained from the National Climate Database of the National Institute of Water and Atmospheric Research (NIWA). The database has been collecting national, regional, and local climate data to serve New Zealand and the world for decades. Data from 72 long-term climate monitoring sites around the country have been used to develop this database; we have made use of the average daily temperature recordings across North Island sites.

During the past century New Zealand's climate has changed significantly as our atmosphere has been affected by greenhouse gases. But while some parts of the world have been getting steadily warmer, New Zealand has instead experienced a number of temperature spikes.

This study's main aim is to sift through and scrutinise the temperature data from the North Island sites to uncover clusters of like kind among them.

**Data Types and Structure**

The dataset includes several key variables:

-   **Site:** The name of the site where the temperature was measured (character).
-   **Date:** The date when the temperature was measured (Date).
-   **Statistic:** The type of temperature measurement (e.g., "Maximum," "Minimum," "Average") (character).
-   **Temperature:** The measured temperature in degrees Celsius (numeric).
-   **Latitude/Longitude (Lat/Lon):** The geographical coordinates of the measurement site (numeric).

This data is structured primarily as a time series, covering multiple decades. Latitude and longitude coordinates enable geographical analysis while temporal information (date, year, month) allows us to do a more in-depth analysis

We made a dataset for analysis that included all the temperature data we had. At first, these data existed in three separate CSV files. We used R to combine the files into one using the full_join function. We were careful not to lose any records or to ignore any data during the merging process. The results preserved the integrity of the time series data across all 30 sites. No missing values were present in the final dataset we used.

## Ethics, Privacy and Security

The data for this project come from public temperature recordings made at climate stations across New Zealand. Because the dataset does not contain personal information and does not involve research on human subjects, the ethical concerns associated with using it are negligible.

It is still essential, though, to use the data with respect and in a manner consistent with its original purpose. We aim to use it to gain a better grasp of climate trends in New Zealand and potentially to help alleviate any misunderstandings about the local effects of climate change. We intend to communicate what we find in a manner consistent with the principles of responsible research - namely, with transparency and accuracy at the forefront.

**Māori Data Sovereignty**

Māori data sovereignty refers to the rights of Māori to control the collection, ownership, and use of any data about them or that relates to their interests. While the temperature data used in this study is not directly about the iwi of these locations, it is data collected about the environment of Aotearoa, New Zealand, meaning it falls under the umbrella of Māori data.

We are gathering findings that could be impactful for Māori, especially concerning the centrality of land and environmental issues to Māori culture and well-being. We are committed to ensuring that this research is respectful of Māori data rights and that it operates within the principles of kaitiakitanga (guardianship) and rangatiratanga (self-determination).

**Privacy and Security Considerations**

Though this research does not make use of sensitive or personal data, it is still critical to maintain the integrity of the dataset. Using version control supports this goal; by keeping a clear record of all the dataset's modifications, we can ensure any unauthorized changes can be easily discovered.

In the event that we do need to discover any "who," "what," or "when," we can do so through GitHub, allowing for a fairly traceable path to any changes made. We can also reverse any changes and uphold the dataset's integrity.

\newpage

## Exploratory Data Analysis

```{r,echo=FALSE}
# Define the North Island sites
north_island_sites <- c(
  "Wellington (Wellington)", "Napier (Hawke's Bay)", "Gisborne (Gisborne)",
  "Tauranga (Bay of Plenty)", "New Plymouth (Taranaki)", "Taumarunui (Manawatū-Whanganui)",
  "Dannevirke (Manawatū-Whanganui)", "Auckland (Auckland)", "Whangārei (Northland)",
  "Hamilton (Waikato)", "Masterton (Wellington)", "Rotorua (Bay of Plenty)",
  "Taupō (Waikato)", "Waiouru (Manawatū-Whanganui)", "Whanganui (Manawatū-Whanganui)",
  "Whangaparāoa (Auckland)"
)
```

```{r, error=TRUE}
library(dplyr)
library(ggplot2) 
library(sf)
library(ggspatial)
```

```{r, error=TRUE}
set.seed(123) clusters <- kmeans(temp$temperature, centers = 4) temp$cluster <- as.factor(clusters$cluster)

temp_north_island <- temp %>% filter(site %in% north_island_sites)
```
## Summary Statistics

```{r, error=TRUE}
summary_stats <- temp_north_island %>% group_by(statistic) %>% summarise( mean_temp = mean(temperature, na.rm = TRUE), median_temp = median(temperature, na.rm = TRUE), sd_temp = sd(temperature, na.rm = TRUE), min_temp = min(temperature, na.rm = TRUE), max_temp = max(temperature, na.rm = TRUE), q1_temp = quantile(temperature, 0.25, na.rm = TRUE), q3_temp = quantile(temperature, 0.75, na.rm = TRUE) ) print(summary_stats)
```
### Average Temperature: 
* Mean(13.52°C): the average temperature across all locations is around 13.5°C. This suggests a relatively mild climate on average. 
* Median(13.5°C): the median matches the mean, indicating a fairly symmetrical data distribution. 
* Standard Deviation(4.36°C): the variation in average temperatures is moderate, showing that most daily averages hover around the mean. 
* Minimum(3.2°C): Some areas or days experience very cold temperatures, though these are rare. 
* Maximum(29.3°C): The highest recorded average temperature approaches 30°C, indicating warmer periods. 
* First and Third Quantiles(10.4°C, 16.8°C): The IQR of 6.4°C shows the spread of central 50% of temperatures implying moderate temperature variation.

### Maximum Temperature: 
* Mean(17.95°C): the average maximum temperature is close to 18°C, which is typical of warmer daytime conditions. 
* Median(17.7°C): similar to the mean, suggesting that extreme values don’t heavily skew the distribution. 
* Standard Deviation(4.69°C): there is a larger spread in maximum temperatures, which might reflect greater daily variability in warm periods. 
* Minimum(-0.9°C): this indicates that even the maximum temperatures occasionally dip below freezing in rare circumstances. 
* Maximum(38°C): some locations experience significant heat, with temperatures soaring up to 38°C, reflecting potential heat waves. 
* First and Third Quantiles(14.6°C, 21.4°C): The IQR of 6.8°C suggests a broad range of daily maximum temperatures between 14.6°C and 21.4°C, covering cooler and warmer days.

### Minimum temperatures: 
* Mean(9.08°C): the average minimum temperature is around 9°C, which indicates cooler nights but not generally extreme cold. 
* Median(9.4°C): similar to the mean, indicating a balanced distribution of minimum temperatures. 
* Standard Deviation(4.86°C): there is a higher variability in minimum temperatures, suggesting that night time temperatures can vary more widely compared to daily averages or maximums. 
* Minimum(-10.1°C): the coldest recorded minimum is very low, possibly from colder, high altitude, or southern regions. 
* Maximum(25.6°C): the highest recorded minimum temperature is 25.6°C, showing that even the nights can be very warm in some instances. 
* First and Third Quantiles(5.8°C, 12.6°C): The IQR of 6.8°C shows that 50% of minimum temperatures fall between 5.8°C and 12.6°C, indicating that many nights fall in their school to mild range.

### General Observations: 
* Symmetry: For both average and maximum temperatures, the mean and median are close, which suggests that these variables are fairly symmetrically distributed. For minimum temperatures, there is a slight skew towards colder values. 
* Range: The wide range between the minimum and maximum temperatures across all types of measurements reflects the diverse climate conditions across North Island locations, from cooler inland regions to warmer coastal areas. 
* Temperature Spread: The larger spread in standard deviations and IQRs for maximum and minimum temperatures suggests greater daily variability, particularly during extreme weather events. * Mild Climate: The data confirms that North Island generally experiences a mild climate, with average temperatures staying within comfortable ranges, but extremes highlight occasional weather volatility.

## Missing Data

```{r, error=TRUE}
missing_data_summary <- temp_north_island %>% group_by(site) %>%
summarise( missing_count = sum(is.na(temperature)), total_obs = n(), missing_percentage = (missing_count/total_obs) * 100 ) %>% arrange(desc(missing_count)) print(missing_data_summary)
```
There are no missing values in the dataset.

## Geographical Analysis:

### Geographical Plots

```{r, error=TRUE}
north_island_sf <- st_as_sf(temp_north_island, coords = c("lon", "lat"), crs = 4326) 
ggplot() + borders("nz", color = "gray85", fill = "gray80") + 
geom_sf(data = north_island_sf, aes(color = temperature), size = 2, alpha = 0.7) + 
scale_color_viridis_c(option = "C", name = "Temperature (Celsius)", limits = c(min(temp_north_island$temperature, na.rm = TRUE), max(temp_north_island$temperature, na.rm = TRUE))) + 
coord_sf(xlim = c(172, 179), ylim = c(-41.5, -34), expand = FALSE) + theme_minimal() + 
labs(title = "Average Temperature at North Island Locations", x = "Longitude", y = "Latitude") + 
annotation_scale(location = "bl", width_hint = 0.5) + 
annotation_north_arrow(location = "tl", which_north = "true", style = north_arrow_fancy_orienteering)
```

Temperature distribution: The plot shows a fairly uniform distribution of average temperatures across North Island, with most locations showing average temperatures in a similar range(between 10°C and 20°C). 
* Higher temperatures: locations near the northernmost part of the island seemed to have a slightly higher average temperature compared to southern locations. 
* Lower temperatures: as remove further South, suggest in Wellington, temperatures appear to decrease slightly indicating cooler conditions.

Geographical Gradient: There is a potential north-south temperature gradient, where temperatures gradually decrease from northern regions(closer to the tropics) to the southern regions(closer to Wellington). This aligns with the geographical expectation that area is closer to the equator tend to be warmer.

No Extreme Temperature Outliers: The temperature range in this plot seems to be fairly moderate. No extreme anomalies such as very high or very low temperatures are immediately visible, suggesting stable climatic conditions across the locations shown.

Coastal vs Inland: * Although difficult to assess without more data, coastal locations suggest Auckland, Tauranga, and Wellington may show more moderate temperatures due to the tempering effect of the ocean. Inland locations like Taupo and Rotorua might exhibit a slightly wider temperature variation but this is not immediately apparent from this plot.

To summarize, this plot confirms the expected temperature patterns in New Zealand North Island, which slightly warmer temperatures in the north and cooler temperatures as you
move South. However it must be noted that no extreme outliers or dramatic geographical temperature variations are immediately noticeable in this representation.

### Visualize Clusters

```{r, error=TRUE}
ggplot() + 
borders("nz", color = "gray85", fill = "gray80") + 
geom_sf(data = north_island_sf, aes(color = as.factor(cluster)), size = 3, alpha = 0.8) + scale_color_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728"), name = "Cluster") + 
coord_sf(xlim = c(170, 180), ylim = c(-41.5, -34), expand = FALSE) + theme_minimal() + 
labs(title = "Geographical Clusters of Temperature Data", x = "Longitude", y = "Latitude") + 
annotation_scale(location = "bl", width_hint = 0.5) + 
annotation_north_arrow(location = "tl", which_north = "true", style = north_arrow_fancy_orienteering)
```

Cluster Distribution: 
* The Cluster 4(red) dominates the southern regions, which suggests that temperature data in southern areas may be more homogeneous or potentially cooler while northern regions may have distinct temperature patterns this is consistent with the earlier plot. 
* The regions are more diverse in their clustering, with clusters 2, 3 and 4 scattered throughout. This could imply that the central part of the island experience is more variability in temperature.

Climate Zones: Cluster 4 (red): likely reflects cooler climates, as it includes Wellington and other southern locations, which are expected to have lower temperatures. Cluster 3 (green) and Cluster 2 (orange) may represent transitional zones with intermediate climate characteristics especially when compared with cluster 4. This clustering suggests distinct temperature patterns based on geography, with the Inland regions showing more variability in temperature, while the southern part of the island has more homogeneity.

### Anomaly Detection:

```{r, error=TRUE}
outlier_detection <- temp_north_island %>% 
filter(statistic == "Average") %>% 
summarise( Q1 = quantile(temperature, 0.25, na.rm = TRUE),
           Q3 = quantile(temperature, 0.75, na.rm = TRUE), 
          IQR = IQR(temperature, na.rm = TRUE) ) 

lower_bound <- outlier_detection$Q1 - 1.5 * outlier_detection$IQR 
upper_bound <- outlier_detection$Q3 + 1.5 * outlier_detection$IQR 

temp_north_island <- temp_north_island %>% 
mutate( outlier = ifelse(statistic == "Average" & (temperature < lower_bound | temperature > upper_bound), "Yes", "No")) 

ggplot(temp_north_island %>% 
filter(statistic == "Average"), aes(x = date, y = temperature, color = outlier)) + 
geom_point(size = 1.0, alpha = 0.5) + 
scale_color_manual(values = c("No" = "cyan", "Yes" = "red"), name = "Outlier") + 
labs(title = "Scatter Plot of Average Temperature with Outliers", y = "Temperature (Celsius)", x = "Date") + 
theme_minimal() + 
theme(legend.position = "bottom")
```

Outlier Distribution: 
* The outliers are marked in red, while the non-outliers are shown in blue. 
* Outliers are concentrated on both the upper and lower extremes of the temperature range specifically: High temperature outliers(above ~ 25°C) are scattered more frequently in the earlier years around pre 1980s, but continue to be observed irregularly overtime. Low temperature outliers(below ~ 5°C) are more common and consistent across the entire timeline, particularly in the latter half of the plot which is post 1950s.

Outlier Trends: 
* A noticeable increase in temperature outliers is seen from around 1960 onward, especially near both the upper and lower boundaries of the temperature range. 
* The presence of outliers near 0°C in the latter half of the timeline could indicate extreme cold events, which may be tied to specific weather patterns or seasonal variations. 
* Extreme high temperatures above 25°C, are more noticeable in the earlier part of the data set around the pre 1970s, although they become less frequent as time progresses.

Time Periods of Interest: 
* There appears to be more concentration of outliers in the early 20th century and toward the present day (post 1980s). This this shift in weather patterns could suggest the changes caused by climate change. Seasonal influences may also play a role, with cold or hot normally is occurring at different periods in the year
This plot highlight significant variations at both temperature extremes, indicating periods of abnormal weather or data anomalies. Further analysis could be done on a much more broader data set, however, within the constraints on the data this is all that’s been concluded.

### Temporal Anomalies

```{r, error = TRUE}
temp_north_island <- temp_north_island %>% 
arrange(date) %>% 
group_by(site) %>% 
mutate(temp_change = temperature - lag(temperature)) 

ggplot(temp_north_island, aes(x = temp_change)) + 
geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) + 
labs(title = "Distribution of Daily Temperature Changes", x = "Temperature Change (Celsius)", y = "Frequency") + 
theme_minimal()
```

Insights from the Histogram: 
* The large peak at around 4°C indicates that most of the daily temperature changes are around this value. This suggests that moderate warming occurs frequently with daily temperatures often increasing by approximately 4°C. 
* The smaller B cat -6°C indicates that there are fewer instances of daily temperature decreases around this value. It’s just that while less common significant daily drops in temperature do occur. 
* The broad spread of temperature changes from -20°C to 15°C indicates a wide range of temperature fluctuations. This suggests variability in temperature changes, with very large increases and decreases occurring although it happens less frequently.
* a left skew in the histogram means that there are more frequent smaller temperature increases, with the tail extended towards larger negative changes. This implies that there are more moderate daily increases in temperature and fewer extreme temperature drops. 
* The plot does not appear to be symmetrical. The lack of symmetry and the presence of a peak around 4°C suggests that the distribution of daily temperature changes is not balanced. Instead there is a tendency towards positive changes(warming) with future currencies of large negative changes(cooling). 
To summarize this, the Lord shows daily temperature changes tend to be moderate with a general tendency towards warming. However there are occasional significant temperature drops and the variability in changes suggest a climate with potential for warming and cooling extremes although pointed more in the warming direction.

### Histograms

```{r north-island-histogram, echo=FALSE, fig.width=14, fig.height=10, fig.cap="Histogram of average temperatures for North Island sites across all seasons, using a binwidth of 1, Each subplot represents one site for visual comparison of temperature distribution"}
source("App.R")

library(ggplot2)



#filter the data for the North Island sites and all seasons
filtered_data <- temp %>%
  filter(site %in% north_island_sites)

p <- ggplot(filtered_data, aes(x = temperature)) + 
  theme_minimal() +
  geom_histogram(binwidth = 1, fill = "blue", alpha = 0.7) +
  ggtitle("Histogram of Average Temperatures", 
          subtitle = "For North Island Sites Across All Seasons") +
  labs(x = "Temperature (Celcius)", y = "Count") +
  facet_wrap(~site, scales = "free", ncol = 4) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    strip.text = element_text(size = 10)
  )

ggsave("North_Island_Histogram_All_Seasons.jpg", plot = p, width = 14, height = 10)


print(p)
```

\newpage

##### **Observations:**

**Shape and Symmetry:** Temperature distributions across many sites such as Auckland (Auckland), Hamilton (Waikato), and Wellington (Wellington) exhibit bell-shaped curves indicative of normal distributions. This data indicates that temperatures tend to cluster around a central mean, with two equally likely extreme temperature distributions on either side of it. Waiouru in Manawatu-Whanganui and Rotorua in Bay of Plenty deviate slightly from this pattern in their distribution patterns; here the distributions appear somewhat irregular. Waiouru shows mild negative skewness (left skew), with longer tails on lower temperature sides; Rotorua shows slight positive skewness (right skew), with longer tails at higher temperature sides. Though these levels of skewness aren't drastic, they suggest these sites experience more frequent extreme temperatures on one end of their temperature spectrum than another.

**Temperature range:** For most sites, the temperature range is from around 0°C to 30°C. However, Waiouru (in Manawatū-Whanganui) and Taumarunui (also in Manawatū-Whanganui) have distributions that extend well into the negative range, which means they occasionally experience average temperatures below freezing, most likely in the winter months. This indicates that these two areas are much more likely to experience extremes on the cold side of the spectrum compared to other locations on the North Island.

Most locations on New Zealand's North Island have average temperatures that fit the usual distribution. However, a few places, such as Waiouru and Taumarunui, are much colder and don't resemble the rest of the North Island. So these two spots and a couple of others help show that there's real regional climatic diversity across the North Island, with some places having quite different conditions from others.

\newpage

### Boxplots per season

Examining the average temperature distributions allows for the construction of an understanding of the climate dynamics of the North Island of New Zealand. Below are four boxplots that represent the four seasons - Spring, Summer, Autumn, and Winter - for all sites on the North Island over a roughly 60-year span. The plots clearly show how the seasonal average temperatures are distributed across the North Island, in addition to what is happening at each site in terms of temperature variance from season to season.

```{r seasonal-boxplots-2x2, echo=FALSE, fig.width=14, fig.height=10, fig.cap="Boxplots of average temperatures for North Island sites across all seasons."}
library(ggplot2)
library(dplyr)
library(gridExtra)

#filter data for each season
summer_data <- temp %>% filter(season == "Summer", statistic == "Average", site %in% north_island_sites)
autumn_data <- temp %>% filter(season == "Autumn", statistic == "Average", site %in% north_island_sites)
winter_data <- temp %>% filter(season == "Winter", statistic == "Average", site %in% north_island_sites)
spring_data <- temp %>% filter(season == "Spring", statistic == "Average", site %in% north_island_sites)

#create individual boxplots
p1 <- ggplot(summer_data, aes(x = factor(site), y = temperature)) +
  geom_boxplot() +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  ) +
  labs(
    title = "Summer Season",
    x = "Site",
    y = "Temperature (Celsius)"
  )

p2 <- ggplot(autumn_data, aes(x = factor(site), y = temperature)) +
  geom_boxplot() +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  ) +
  labs(
    title = "Autumn Season",
    x = "Site",
    y = "Temperature (Celsius)"
  )

p3 <- ggplot(winter_data, aes(x = factor(site), y = temperature)) +
  geom_boxplot() +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  ) +
  labs(
    title = "Winter Season",
    x = "Site",
    y = "Temperature (Celsius)"
  )

p4 <- ggplot(spring_data, aes(x = factor(site), y = temperature)) +
  geom_boxplot() +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  ) +
  labs(
    title = "Spring Season",
    x = "Site",
    y = "Temperature (Celsius)"
  )

#arrange the plots in a 2x2 grid
grid.arrange(p1, p2, p3, p4, ncol = 2, nrow = 2)


```

##### **Summer:**

The highest median temperatures, normally between 15ºC and 22ºC, are found in the summer. In this season, places such as the coastal Northland city of Whangarei and the Bay of Plenty city of Tauranga enjoy their median warmest temperatures. On the other hand, inland summer sites such as Waiouru (Manawatū-Whanganui) do not come close to the kinds of temperatures experienced at these northern coastal towns.

In some respects, summer is also the most predictable of New Zealand's seasons, with its warm temperatures reliably extending over prolonged periods. This is reflected by summer having the narrowest IQR. However, when it comes to the temperature extremes, the IQR and the median in the warmest summer location do not appear to be significantly higher than the temperature spread and median in places that have their warmest median in spring.

##### **Autumn:**

Autumn serves as an indicator of winter's approach, with median temperatures generally falling from the high summer temperatures of 10ºC-15ºC across all sites and coastal regions experiencing slightly higher median temps compared to inland locations - this seasonal shift provides clear warning of its approach.

Autumn temperatures vary with each passing season and its variability affects temperature IQR more than winter IQR; Gisborne (Gisborne) and New Plymouth (Taranaki) see narrower IQRs suggesting more stable temperatures this autumn compared with winter; Waiouru in Manawatu-Whanganui still experiences some lower outliers below 5ºC due to early frosts or cooler nights typical of this transitional time of year.

##### **Winter:**

The winter months, when median temperatures are between 5°C and 12°C, are the coldest in most parts of the North Island. Areas such as Taupo (Waikato) and Waiouru (Manawatū-Whanganui) are some of the coldest places in the North Island during this time, hitting median temperatures closer to 5°C. The IQR for this season is the narrowest of any season, indicating that it is mostly stable but mostly cold. The IQR is slightly lower in Auckland (Auckland) and Whangarei (Northland) because these coastal areas are likely experiencing the milder conditions associated with the ocean's moderating effect. We noted above that Waiouru has significant outliers that hit below 0°C, implying these numbers reflect a few times when it was freezing cold across these sites - a common occurrence in this region of the North Island.

##### **Spring:**

Spring is a time of shifting weather patterns and a moderate temperature range across the North Island. Generally, spring temperatures range from 12°C to 18°C. For the most part, inland areas like Taupo and Waiouru are cooler than coastal areas like Auckland and Whangarei. The ocean's moderating effect keeps coastal temperatures milder, with fewer ups and downs, than those experienced in wider Consistent with their average temperatures, inland areas do tend to have moderate ranges.

The North Island sites reveal substantial temperature differences that reflect the typical expected seasonal patterns in New Zealand. The inland and more elevated sites - especially Waiouru (in the Manawatū-Whanganui region) and Taupo (in the Waikato region) - show lower average median temperatures. They also have a wider range of temperatures, particularly in winter and spring. In contrast, the coastal sites have average temperatures that are a little more stable and a little less warm than those of the inland and coastal sites.

\newpage

### Correlation Matrix

```{r site-correlation-matrix, echo=FALSE, fig.width=14, fig.height=10, fig.cap="Correlation Matrix of Average Temperatures by Site."}

library(tidyr)

#filter the data for 'Average' statistic and aggregate over all seasons
average_data <- temp %>%
  filter(statistic == "Average", site %in% north_island_sites) %>%
  group_by(site) %>%
  summarise(mean_temp = mean(temperature, na.rm = TRUE)) %>%
  ungroup()

#pivot data to wide format, where each site becomes a column
wide_data <- temp %>%
  filter(statistic == "Average", site %in% north_island_sites) %>%
  select(site, date, temperature) %>%
  group_by(site, date) %>%
  summarise(mean_temp = mean(temperature, na.rm = TRUE)) %>%
  pivot_wider(names_from = site, values_from = mean_temp)

cor_matrix <- cor(wide_data[-1], use = "complete.obs")

#convert correlation matrix to long format for plotting
cor_data <- as.data.frame(as.table(cor_matrix))
cor_data$Var1 <- as.factor(cor_data$Var1)
cor_data$Var2 <- as.factor(cor_data$Var2)

p <- ggplot(cor_data, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Freq, 2), color = ifelse(Freq == 1, "white", "black")), size = 3) + # Conditional text color
  scale_color_identity() + # Use the color identity scale
  scale_fill_gradientn(colors = RColorBrewer::brewer.pal(11, "RdBu"), na.value = "grey90") +
  theme_minimal() +
  labs(
    x = "Site",
    y = "Site",
    fill = "Correlation",
    title = "Correlation Matrix of Average Temperatures by Site"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


p

```

**Observations:** The average temperature correlation matrix for North Island sites provides some clear signals about the temperature relationships among these locations. Most of the correlation coefficients fall within a rather narrowly defined range of 0.85 to 1.00. The sites' temperatures tend to move in tandem, indicating that they're under the influence of the same (or very similar) set of climatic factors and yielding a pretty well-defined and cohesive temperature pattern across the North Island.

However, there are some notable deviations. One of them is Wellington, which has somewhat lower correlations with other sites, sometimes dipping as low as 0.85. This might reflect the way Wellington's geography, with its urban topography and coastal-wind conditions, partly determines a rather special microclimate. Another site with lower correlations is Gisborne; its location on the east coast, with its somewhat sheltered situation, probably also allows it to indulge in a special microclimate. On the other hand, locations in the central North Island—such as Hamilton (Waikato), Rotorua (Bay of Plenty), and Taupō (Waikato)—show very strong correlations that are often near 1.00. The sites have similar environmental conditions, which accounts for the similarity of their temperature patterns.

In general, this matrix of correlations highlights the vigorous temperature relationships at the majority of North Island sites. A few exceptions can be attributed to geographical and climatic factors that provide effective explanations for them.

\newpage

### Time Series Plots

```{r time-series-plot, echo=FALSE, fig.width=14, fig.height=10, fig.cap="Temperature Over Time (90-day smoothing) by Site."}

#subsetting to a representative list of locations to make plots more readable
subset <- list("Whangārei (Northland)", "Auckland (Auckland)", "Taupō (Waikato)", "Gisborne (Gisborne)", "New Plymouth (Taranaki)", "Wellington (Wellington)")

ts_data <- temp %>%
  filter(statistic == "Average", site %in% north_island_sites, date > 1940-01-01) %>%
  filter(site %in% subset)


#calculate the 90-day rolling average
smooth <- ts_data %>%
  group_by(site) %>%
  mutate(rolling_avg = rollmean(temperature, k = 90, fill = NA, align = "right")) %>%
  ungroup()

#plot the time series data with 90-day smoothing
p <- ggplot(smooth, aes(x = date, y = rolling_avg, group = site, color = site)) +
  geom_line(size = 0.5) +
  facet_wrap(~ site, ncol = 2) +
  labs(
    title = "Temperature Over Time Since 1940 (90-day smoothing)",
    x = "Year",
    y = "Temperature (\u00B0C)"
  ) +
  stat_smooth(colour = "red", linewidth=0.5) + 
  theme(legend.position = "none",
    strip.text = element_text(size = 10)
  )

print(p)
```

\newpage

```{r seasonal-plot, echo=FALSE, fig.width=14, fig.height=10, fig.cap="Seasonal Plots"}

monthly <- ts_data %>%
  group_by(site, month_year) %>%
  summarise(mean_temp = mean(temperature, na.rm = TRUE)) %>%
  ungroup()

monthly_subset <- monthly %>%
  filter(site %in% subset)

monthly_subset %>% 
  mutate(month_year = yearmonth(as.character(month_year))) %>%
  as_tsibble(key = site, index=month_year) %>%
  gg_season(mean_temp) +
  facet_wrap(~ site, ncol = 2) +
  labs(
    title = "Monthly Temperature Trend Since 1940",
    x = "Year",
    y = "Temperature (\u00B0C)"
  )
  
```

**Observations:**

The time series data show the clear (and expected) seasonal pattern in average temperature.

The trend lines in the temperature over time plots show gentle fluctuations in average temperature over time and a small overall increase. This increase is more obvious in the seasonal plot, where the pink lines indicating the past decade clearly sit above those lines indicating earlier years.

The seasonal plot also shows that although the various locations experience occasional months with warmer or cooler average temperatures, by and large the temperature follows the seasonal pattern each year.

\newpage

### Individual Contributions

##### Arman

-   **Histogram Analysis**: The histogram plots were designed, implemented, and analysed by Arman. He filtered the data, determined the bin widths, and wrote the analysis section that discusses the patterns and distributions observed in the data.

-   **Boxplot Analysis**: The boxplot visualisations for the different seasons were created by Arman. He did the coding for the plots, made sure they represented temperature ranges across the sites appropriately, and interpreted the why of the seasonal differences. He did the detailed write-up of the boxplots, which highlight some of the important findings regarding temperature trends across different seasons.

-   **Correlation Matrix Analysis**: Arman created a correlation matrix to explore relationships among temperature variations at various sites across North Island. He managed data processing, matrix calculation, visualisation and provided written analyses on temperature correlations among them.

-   **General Contributions**: In addition to the specific duties noted above, Arman contributed to the Background and Data and the Ethics, Privacy, and Security sections of the report. These sections were developed collaboratively, with each team member providing input to achieve the necessary accuracy and completeness in explaining the data source, structure, and ethical aspects of our work.

##### Amanda

-   **Time Series**: Amanda developed the time series plots and provided the written analysis.
-   **GitHub**: Amanda set up the group project on GitHub

##### Shubham

- **Overview of data**: Shubham did the basic overview providing the basic summary statistics for the temperature data, including mean, median, standard deviation, minimum, maximum, and quartiles for each type of measurement (min, max, and average temperatures) across all sites.

- **Geographical analysis**: Shubham did the geographic plots by overlaying the temperature data onto the North Island to identify any patterns. Also made visualization for the clusters to check for any geographical significance.

- **Anomaly Detection**: Made a histogram to check for temporal anomalies that can indicate data anomalies or extreme weather events.

